### 2016-11-11 Paper_v1_.tex
Nov11_4

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Toward a Genome Scale Dynamic Model of Cell-Free Protein Synthesis in \emph{Escherichia~coli}}}}
\vspace{0.05in}
{\par \centering \large{Nicholas Horvath, Michael Vilkhovoy, Joseph Wayman, Kara Calhoun$^{1}$, James Swartz$^{1}$ and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {Robert Frederick Smith School of Chemical and Biomolecular Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering {$^{1}$School of Chemical Engineering}}
{\par \centering {Stanford University, Stanford, CA 94305}}
{\par \centering \textbf{Running Title:}~Dynamic modeling of cell-free protein synthesis}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Scientific~Reports}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, Robert Frederick Smith School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Cell-free protein expression systems have become widely used in systems and synthetic biology.
In this study, we developed an ensemble of dynamic \textit{E. coli} cell-free protein synthesis (CFPS) models.
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble described the training data, with the exception of some of the amino acid dynamics.
To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
balance analysis. The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle,
while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as
GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

@@

In this study, we developed an ensemble of \textit{E. coli} cell-free protein synthesis (CFPS) models using the hybrid cell-free modeling approach of Wayman et al [REFHERE].
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble described the training data, with the exception of some of the amino acid dynamics.
To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
balance analysis. The observed CAT yield was 25\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle,
while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as
GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
% Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
% We integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
% This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

@@

\subsection*{Generation of model ensemble}
We generated an ensemble of 100 diverse parameter sets via a Markov chain Monte Carlo random walk.
Beginning with a single parameter set as a starting point, we calculated its cost function, equal to the sum-squared-error between experimental data and model predictions:
\begin{equation}\label{eqn:cost-function}
    cost=\sum_{i=1}^{\mathcal{D}}\left(w_i\sum_{j=1}^{\mathcal{T}_i}abs\left(x_{ij}^{data}-x_{i}^{sim}|_{t(j)}\right)\right)
\end{equation}
where $\mathcal{D}$ denotes the number of datasets, $w_i$ denotes the weight of the $i$th dataset, $\mathcal{T}_i$ denotes the number of timepoints in the $i$th dataset, $t(j)$ denotes the $j$th timepoint, $x_{ij}^{data}$ denotes the value of the $i$th dataset at the $j$th timepoint, and $x_{i}^{sim}|_{t(j)}$ dneotes the simulated value of the metabolite corresponding to the $i$th dataset, interpolated to the $j$th timepoint.
We then perturbed model parameters:
\begin{equation}\label{eqn:parameter-perturbation}
    k_i^{new}=k_i*exp(a\,r_i)\qquad{i=1,2,\hdots,\mathcal{P}}
\end{equation}
where $\mathcal{P}$ denotes the number of parameters, equal to 815, which includes 163 rate constants, 163 enzyme degradation rate constants, 455 saturation constants, and 34 control parameters, $k_i^{new}$ denotes the new value of the $i$th parameter, $k_i$ denotes the current value of the $i$th parameter, $a$ denotes a distribution variance, and $r_i$ denotes a random sample from the normal distribution.
For each newly generated parameter set, we re-solved the balance equations and calculated the cost function.
All sets with a lower cost than the previous set, and some with higher cost, were added to the ensemble.
After generating 12,437 sets, we selected 100 sets with minimal correlation to each other so as to avoid over-sampling any region of parameter space.
The original 12,437-set ensemble had a mean Pearson correlation coefficient [REFERENCE NEEDED?] of 0.93 between pairs of sets; the 100-set ensemble had a mean Pearson correlation coefficient of 0.84 between pairs of sets.

\subsection*{Sensitivity analysis}
We determined the reactions most important to protein production by computing the local sensitivity of CAT concentration to each rate constant in the network.
Each sensitivity index was formulated as:
\begin{equation}\label{eqn:local-sensitivity}
\begin{split}
    S_{ij}=norm(CAT(p_i,p_j,t)-CAT(\alpha*p_i,\alpha*p_j,t))\qquad{i,j=1,2,\hdots\mathcal{P}}\\
\end{split}
\end{equation}
where $S_{ij}$ denotes the sensitivity of CAT production to the $i$th and $j$th parameters, $CAT(p_i,p_j,t)$ denotes CAT concentration as a function of time and the $i$th and $j$th parameters, $\alpha$ denotes the perturbation factor, equal to 1.01, and $\mathcal{P}$ denotes the number of rate constants, equal to 163.
In calculating the pairwise sensitivities, each parameter was perturbed by 1\%; first-order sensitivities ($i$ = $j$) were subject to two 1\% perturbations, equivalent to a perturbation of 2.01\%.

@@

\begin{figure}[ht]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/CarbonBoth.pdf}
\caption{Central carbon metabolism in the presence (top) and absence (bottom) of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species. Best-fit parameter set (orange line) versus experimental data (points). 95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
\label{fig:CarbonBoth}
\end{figure}

>>>>

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Toward a Genome Scale Dynamic Model of Cell-Free Protein Synthesis in \emph{Escherichia~coli}}}}
\vspace{0.05in}
{\par \centering \large{Nicholas Horvath, Michael Vilkhovoy, Joseph Wayman, Kara Calhoun$^{1}$, James Swartz$^{1}$ and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {Robert Frederick Smith School of Chemical and Biomolecular Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering {$^{1}$School of Chemical Engineering}}
{\par \centering {Stanford University, Stanford, CA 94305}}
{\par \centering \textbf{Running Title:}~Dynamic modeling of cell-free protein synthesis}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Scientific~Reports}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, Robert Frederick Smith School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Cell-free protein expression systems have become widely used in systems and synthetic biology.
In this study, we developed an ensemble of dynamic \textit{E. coli} cell-free protein synthesis (CFPS) models.
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble described all of the training data, especially the central carbon metabolism.
To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux balance analysis.
The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
%The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle, while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

@@

In this study, we developed an ensemble of \textit{E. coli} cell-free protein synthesis (CFPS) models using the hybrid cell-free modeling approach of Wayman et al [REFHERE].
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble described the training data, with the exception of some of the amino acid dynamics.
To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield with the maximum theoretical CAT carbon yield calculated using sequence specific flux balance analysis.
The observed CAT yield was 25\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle, while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
% Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
% We integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
% This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

@@

\subsection*{Generation of model ensemble}
We generated an ensemble of 100 diverse parameter sets via a Markov chain Monte Carlo random walk.
Beginning with a single parameter set as a starting point, we calculated its cost function, equal to the sum-squared-error between experimental data and model predictions:
\begin{equation}\label{eqn:cost-function}
    cost=\sum_{i=1}^{\mathcal{D}}\left(w_i\sum_{j=1}^{\mathcal{T}_i}abs\left(x_{ij}^{data}-x_{i}^{sim}|_{t(j)}\right)\right)
\end{equation}
where $\mathcal{D}$ denotes the number of datasets, equal to 37, $w_i$ denotes the weight of the $i$th dataset, $\mathcal{T}_i$ denotes the number of timepoints in the $i$th dataset, $t(j)$ denotes the $j$th timepoint, $x_{ij}^{data}$ denotes the value of the $i$th dataset at the $j$th timepoint, and $x_{i}^{sim}|_{t(j)}$ denotes the simulated value of the metabolite corresponding to the $i$th dataset, interpolated to the $j$th timepoint.
We then perturbed model parameters:
\begin{equation}\label{eqn:parameter-perturbation}
    k_i^{new}=k_i*exp(a\,r_i)\qquad{i=1,2,\hdots,\mathcal{P}}
\end{equation}
where $\mathcal{P}$ denotes the number of parameters, equal to 815, which includes 163 rate constants, 163 enzyme degradation rate constants, 455 saturation constants, and 34 control parameters, $k_i^{new}$ denotes the new value of the $i$th parameter, $k_i$ denotes the current value of the $i$th parameter, $a$ denotes a distribution variance, and $r_i$ denotes a random sample from the normal distribution.
For each newly generated parameter set, we re-solved the balance equations and calculated the cost function.
All sets with a lower cost than the previous set, and some with higher cost, were added to the ensemble.
After generating 12,437 sets, we selected 100 sets with minimal correlation to each other so as to avoid over-sampling any region of parameter space.
The original 12,437-set ensemble had a mean Pearson correlation coefficient [REFERENCE NEEDED?] of 0.94 between pairs of sets, while the 100-set ensemble had a mean Pearson correlation coefficient of 0.83 between pairs of sets.

\subsection*{Local sensitivity analysis}
We determined the reactions most important to protein production by computing the local sensitivity of CAT concentration to each rate constant in the network.
Each sensitivity index was formulated as:
\begin{equation}\label{eqn:local-sensitivity}
\begin{split}
    S_{ij}=norm(CAT(p_i,p_j,t)-CAT(\alpha*p_i,\alpha*p_j,t))\qquad{i,j=1,2,\hdots\mathcal{P}}\\
\end{split}
\end{equation}
where $S_{ij}$ denotes the sensitivity of CAT production to the $i$th and $j$th parameters, $CAT(p_i,p_j,t)$ denotes CAT concentration as a function of time and the $i$th and $j$th parameters, $\alpha$ denotes the perturbation factor, equal to 1.01, and $\mathcal{P}$ denotes the number of rate constants, equal to 163.
In calculating the pairwise sensitivities, each parameter was perturbed by 1\%; first-order sensitivities ($i$ = $j$) were subject to two 1\% perturbations, equivalent to a perturbation of 2.01\%.
The array of sensitivity indices was organized into a clustergram (Fig.~\ref{fig:Sensitivity}, top) with both axes corresponding to the 163 rate constants rearranged from most sensitive (top, right) to least sensitive (bottom, left).

Likewise, we determined the reactions most important to the system as a whole by computing the local sensitivity of the species for which data exist to each rate constant in the network.
Each sensitivity index was formulated as:
\begin{equation}\label{eqn:local-sensitivity}
\begin{split}
    S_{ij}=norm(X(p_i,p_j,t)-X(\alpha*p_i,\alpha*p_j,t))\qquad{i,j=1,2,\hdots\mathcal{P}}\\
\end{split}
\end{equation}
where $S_{ij}$ denotes the sensitivity of the system state to the $i$th and $j$th parameters, and $X(p_i,p_j,t)$ denotes the system state, an array consisting of the concentration of every species for which data exist as a function of time and the $i$th and $j$th parameters.
% $\alpha$ denotes the perturbation factor, equal to 1.01, and $\mathcal{P}$ denotes the number of rate constants, equal to 163.
This array of sensitivity indices was also organized into a clustergram (Fig.~\ref{fig:Sensitivity}, bottom).

@@

% Figures and captions go here ...
\begin{figure}[ht]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/CarbonBoth.pdf}
\caption{Central carbon metabolism in the presence (top) and absence (bottom) of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species. Best-fit parameter set (orange line) versus experimental data (points). 95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
\label{fig:CarbonBoth}
\end{figure}
